---
title: Assignment 3 Due Friday, October 18 at midnight
author: 'Andrew Kerr'
format:
  html:
    embed-resources: true
  #pdf: default
editor: 
  markdown: 
    wrap: 72
---

Download the .qmd source for this document
[here](https://raw.githubusercontent.com/juliaSchedler/Stat416_CourseWebsite/refs/heads/main/Assignments/Assignment3.qmd).

## 1. Math

Recall the so-called "trend-stationary" model $$
x_t = \mu_t + y_t
$$

1.  \[3 points\] There are three terms (symbols) in that equation. Which
    symbol corresponds to the trend, and which corresponds to the
    stationary part? What would you call the remaining symbol?

    $x_t$: observations

    $\mu_t$: trend

    $y_t$: stationary process

2.  We have considered a few possible models for the trend: $$
    \begin{aligned}
    \mu_{LM,t} &= \beta_0 + \beta_1 t  \\
    \mu_{RW,t} &=  \delta + \mu_{RW, t-1} + w_t
    \end{aligned}
    $$

3.  \[2 points\] What do "RW" and "LM" stand for?

    RW: random walk

    LW: linear model

4.  \[4 points\] Between $\mu_{LM,t}$ and $\mu_{RW,t}$, which has "more
    interesting" temporal structure\[2 points\]? Why\[2 points\]?

    $\mu_{RW, t}$ has a more interesting temporal structure because it
    can have trending, cycles, and unpredictable movements. On the other
    hand, linear models represent a trend that is assumed to be a fixed
    linear relationship over time, making a predictable and stable
    pattern.

5.  \[4 points\] Write down the equations for $x_{LM, t}$ and $x_{RW,t}$

    $x_{LM, t} = \beta_0 + \beta_1t + y_t$

    $x_{RW,t} = \delta + x_{t-1} + w_t$:

Consider this (edited) excerpt from the textbook (page 49):

::: {.callout-tip appearance="simple" icon="false"}
## Differencing vs. Detrending

One advantage of differencing over detrending to remove trend is that no
parameters are estimated in the differencing operation. One
disadvantage, however, is that differencing does not yield and estimate
of the stationary process $y_t$.

For example, if we difference a random walk $x_t$ with drift $\delta$,

$$
x_t - x_{t-1} = \delta + w_t + y_t - y_{t-1}.
$$

If an estimate of $y_t$ is essential, then detrending may be more
appropriate. This would be the case, for example, if we were interested
in the business cycle of commodities.

If the goal is to coerce the data to stationarity, then differencing may
be more appropriate. Differencing is also a viable tool if the trend is
fixed, that is, when using $\mu_{LM,t}$ as the trend model, we have: $$
x_t - x_{t-1} = \beta_1 + y_t - y_{t-1}
$$ Because differencing plays a central role in time series analysis, it
receives its own notation. The first difference is denoted: $$
\nabla x_t = x_t - x_{t-1}
$$ As we have seen, the first difference eliminates a linear trend. A
second difference can eliminate a quadratic trend, and so on.
Differences of order $d$ are denoted: $$
\nabla^d = (1-B)^d
$$
:::

5.  \[1 point\] Take a look in the book. What is $B$ called?

    Backshift operator

6.  \[6 points\] Why is it an "advantage" that no parameters are
    estimated in the differencing operation?

    This is an advantage because this reduces the risk of over fitting
    to the data and reduces the complexity compared to detrending.

7.  \[6 points\] Connect part 2 question 9 below to a sentence in the
    above excerpt.

    "...the first difference eliminates a linear trend."

8.  \[10 points\] What kind of plot might you make to check if the data
    has been "coerced to stationarity"?

    A scatter/line plot of the data itself and ACF or ACF1 plot.

9.  \[3 points\] Rate your math anxiety (1 = effortless, 100 =
    nightmare) while working on this problem.

    Questions 1-4: 10

    Questions 5-8: 20

    --\> 15

## 2. Data Analysis (code)

1.  \[10 points\] Adapt the code from Example 1.2 in the book to plot
    just the global land temperature series in a time series plot. \[2
    points\]. Describe the structure of the trend and/or seasonal
    components, if present\[8 points\].

```{r}
#| label: part 1
#| message: false

library(astsa)

culer = c(rgb(.85,.30,.12,.6), rgb(.12,.65,.85,.6))
tsplot(gtemp_land, col=culer[1], lwd=2, type="o", pch=20,
ylab="Temperature Deviations", main="Global Warming for Land Surface")
```

This plot of global land temperatures has an increasing trend,
increasing as time goes on at a larger rate at later times. Although
there are fluctuations year to year, there does not appear to be a
seasonal pattern present.

2.  \[4 points\] How frequently were the observations collected?

```{r}
#| label: part 2
#| message: false

unique(diff(time(gtemp_land)))
```

The observations were collected yearly.

3.  \[5 points\] Plot the autocorrelation function of the global land
    temperature series. Comment on the temporal structure.

```{r}
#| label: part 3
#| message: false

acf(gtemp_land)
```

As time between observations increases, the correlation between the two
points decreases. But, since the ACF value remains above 0 for so long
this means that observations are influenced by past observations for a
long time. Additionally, there is a slight "scalloped" shape, which
suggests a possible seasonal pattern.

4.  \[15 points\] Estimate the trend of the series using a symmetric,
    equally weighted 5-point moving average\[4 points\]. Plot the trend
    estimate on top of the data \[4 points\]. Comment on the trend--
    does it reveal any patterns difficult to see in the data?\[7
    points\]

```{r}
#| label: part 4
#| message: false

w = c(.5, rep(1,3), .5)/5
land_ma = stats::filter(gtemp_land, sides=2, filter=w)
tsplot(gtemp_land, type = 'p', main = paste("Global Land Temperature with 5-Point Moving Average"), xlab = "Year", ylab = "Temperature Deviations")
lines(gtemp_land)
lines(land_ma, col = 'firebrick', lwd = 2)
legend("topleft", legend = c("Original Data", "5-Point Moving Average"), 
       col = c("black", "firebrick"), lty = 1)
```

The moving average smoother helps show the possible cyclic pattern, but
overall continues to show an overall increasing trend that is steeper in
more recent years.

5.  \[10 points\] Estimate the trend of the series using a linear
    regression on time\[3 points\]. Plot the trend estimate on top of
    the data\[3 points\]. Is this trend estimate comparable to the
    moving average?\[4 points\]

```{r}
#| label: part 5
#| message: false

land_lm <- lm(gtemp_land ~ time(gtemp_land))
tsplot(gtemp_land, type = 'p', main = paste("Global Land Temperature with Linear Model"), xlab = "Year", ylab = "Temperature Deviations")
lines(gtemp_land)
abline(land_lm, col = 'firebrick', lwd = 2)
legend("topleft", legend = c("Original Data", "Linear Regression Line"), 
       col = c("black", "firebrick"), lty = 1)
```

The trend shown by the linear model also shows an upward trend. Unlike
the moving average, this trend line cannot show the increase in upward
trend in more recent years since it is a straight line.

6.  \[15 points\] De-trend the data with respect to each of the trends
    you estimated \[4 points each\]. Make a time series plot of each
    result \[1 point each\] and an acf of each result \[1 point each\].
    Comment on the temporal structure in each of the plots.

```{r}
#| label: part 6
#| message: false

par(mfrow = c(2,2))

land_detrend_ma <- gtemp_land - land_ma
tsplot(land_detrend_ma, type = 'p', main = paste("Detrended 5-Point Moving Average"), xlab = "Year", ylab = "Detrended Temperature Deviations")
lines(land_detrend_ma)

land_detrend_lm <- resid(land_lm)
tsplot(land_detrend_lm, type = 'p', main = paste("Detrended Linear Model"), xlab = "Time", ylab = "Residuals")
lines(land_detrend_lm)

invisible(acf1(land_detrend_ma, na.action = na.pass))

invisible(acf1(land_detrend_lm))

par(mfrow = c(1,1))
```

The detrended moving average plot shows a slight increasing trend, while
the detrended linear model plot decreases then increases. The ACF1 plot
for the detrended moving average shows most values are near zero, with
increases at lag 3, 6, 8, and 12 suggesting potential seasonal or
cyclical patterns, but overall removing most of the correlation. On the
other hand, the ACF1 plot of the detrended linear model shows decreasing
correlation as time points move further apart (higher lags), which means
that this model may not have captured all of the patterns in the data.

7.  \[20 points\] Difference the time series and plot the result\[5
    points\]. Also compute the acf and plot it\[5 points\]. Comment on
    whether the differencing has "coerced the data to stationarity"\[10
    points\].

```{r}
#| label: part 7
#| message: false

par(mfrow = c(2,1))

land_diff <- diff(gtemp_land)
tsplot(land_diff, type = 'p', main = paste("Differenced Global Land Temperature"), xlab = "Year", ylab = "Differenced Temperature Deviations")
lines(land_diff)

invisible(acf(land_diff))

par(mfrow = c(1,1))
```

Differencing the data has not coerced it to stationary. Although the
mean appears stationary since it is centered around zero and does not
fluctuate much, the correlation has a large negative spike at lag 2 and
a slight positive spike at lag 12. Therefore, differencing the data has
helped move the time series toward stationarity.

8.  \[11 points\] Estimate the sum of squared error for both the linear
    regression and the moving average \[3 points each\]. What do they
    suggest about which model is "better"?\[3 points\] Does this agree
    with your visual assessments in 3 and 4?\[2 points\]

```{r}
#| label: part 8
#| message: false

sum(as.vector(land_detrend_ma)^2, na.rm = T)
sum(land_detrend_lm^2)
```

Since the SSE for the moving average model is smaller, this is the
better model. This agrees with what I concluded from my visual
assessments.

9.  \[10 points\] Why can't you do part 6 for the differenced series?

    We cannot detrend a differenced series because the first
    differencing removes linear trend, so it has already removed the
    trend that detrending is trying to extract (differencing is a method
    of detredning).

10. \[3 points\] Ensure that when you render the document to .html to
    turn in that you have `embed-resources: true` in the options at the
    top of the document (you may also turn in a pdf). Also, set message
    to false for all the code chunks.

    Okay!

## 3. The literature

The original paper for the data set you just analyzed [can be found
here](https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/JD092iD11p13345).

1.  \[2 points\] Set a timer for 5 minutes. Find one sentence you feel
    you understand, and one you do not understand but would like to
    understand better.

    I understand: "We calculate the mean of both records for the period
    in common, and adjust the entire second record ($T_2$) by the
    difference (bias) $\delta T$."

    I want to understand: "An alternative method commonly used to
    combine station records is to define $\delta T$ by specifying the
    mean temperature of each station as zero for a specific period which
    had a large number of stations, for example 1950-1980; this
    alternative method compares unfavorably to ours with regard to both
    making use of the maximum number of stations and defining the bias
    $\delta T$ between stations as accurately as possible."

2.  \[4 points\] Take a look at Figure 6. There are six time series
    plotted. One is the data plotted in number 1 of the coding portion,
    the other is the moving average estimate estimated in number 3.
    Which of the six series are they?

    They are the two time series overlayed on-top of each other on the
    top of Figure 6 labeled Global. The data is the dotted line while
    the moving average is the solid line.

3.  \[4 points\] The textbook for our class states on page 3 that "the
    data are annual temperature anomalies averaged over the Earth's land
    area.". Does the book state what is specifically meant by "anomaly"
    here?

    The textbook states that the anomalies (values) are "deviations
    (degrees C) from the 1951–1980 average."

4.  \[2 points\] Consider the concept of averaging over the Earth's land
    area, then take a look at Figure 1. Describe in terms of the
    "circles" how you might calculate that global average.

    I might calculate the global average by taking the average
    temperature of each circle, then taking a weighted average of all of
    these values. I would weight them by what percentage of the circle
    is on land vs. over the water. Circles with higher land coverage
    would be weighted more heavily than those with higher water
    coverage. These weights would also have to account for the overlap
    of the circles so one area of the globe does not get too much
    influence (although I am not sure how to do this part exactly, maybe
    penalize the weight by percentage overlap).

5.  \[15 points\] Figure 3 plots correlation coefficients between annual
    mean temperature changes for pairs of randomly selected stations
    having at least 50 common years in their records. How can you
    rephrase "annual mean temperature changes" in terms of detrending or
    differencing?

    We can rephrase this as taking the first differences between the
    annual mean temperature time series at each station.

## 4. Weights

1.  \[6 points\] One might consider the points I have given to each
    numbered problem as "weights". Based on the weights, which content
    is the most important? Answer on both the individual problem level
    and the section level (i.e. math, code, etc).

    The most important question is question 7 from the coding section on
    differencing (20 points), while the most important section is the
    coding section with 103 total points (59% of the points for this
    assignment).
