---
title: "Lecture14"
format: html
editor: visual
---

# Activity 1 - Problem 8.1 in FPP3
Consider the the number of pigs slaughtered in Victoria, available in the aus_livestock
dataset.

a. Use the ETS() function in R to estimate the equivalent model for simple exponential
smoothing. Find the optimal values of 𝛼 and ℓ0, and generate forecasts for the next four
months.

```{r}
library(fpp3)

fitted_pigs <- aus_livestock %>%
  filter(State == 'Victoria', Animal == 'Pigs') %>%
  model(pigs = ETS(Count ~ error('A') + trend('N') + season('N')))

fitted_pigs %>% report()

pred_pigs <- fitted_pigs %>% forecast(h = 4)
```

b. Produce a prediction interval in R.

```{r}
pred_pigs |>
  head(1) |>
  mutate(interval = hilo(Count, 95)) |>
  pull(interval)
```

c. Bonus: Compute a 95% prediction interval for the first forecast using ̂ 𝑦 ± 1.96𝑠 where
𝑠 is the standard deviation of the residuals. Compare your interval with the interval
produced by R.

```{r}
s <- augment(fitted_pigs) |>
pull(.resid) |>
sd()

yhat <- pred_pigs |>
pull(.mean) |>
head(1)
yhat + c(-1, 1) * 1.96 * s
```

# Activity 2 - Problem 8.5 in FPP3
Data set global_economy contains the annual Exports from many countries. Work with data
from Turkey.

a. Plot the Exports series and discuss the main features of the data.

```{r}
library(astsa)

turk_econ <- global_economy %>%
  filter(Country == "Turkey") %>%
  select(Year, Exports)

turk_econ %>% autoplot(Exports)
```

b. Use an ETS(A,N,N) model to forecast the series, and plot the forecasts.

```{r}
fit_ann <- turk_econ %>%
  model(ANN = ETS(Exports ~ error('A') + trend('N') + season('N')))

fit_ann %>% forecast(h = 5) %>% autoplot(turk_econ)
```

c. Compute the RMSE values for the training data.

```{r}
accuracy(fit_ann) %>% pull(RMSE)
```

d. Compare the results to those from an ETS(A,A,N) model. (Remember that the trended
model is using one more parameter than the simpler model.) Discuss the merits of the
two forecasting methods for this data set.

```{r}
fit <- turk_econ %>%
  model(ANN = ETS(Exports ~ error('A') + trend('N') + season('N')),
        AAN = ETS(Exports ~ error("A") + trend("A") + season("N")))
```

e. Compare the forecasts from both methods. Which do you think is best?

```{r}
fit %>% accuracy()
```

f. Calculate a 95% prediction interval for the first forecast for each series, using the RMSE
values and assuming normal errors. Compare your intervals with those produced using
R.

```{r}
fit %>%
  forecast(h = 10) %>%
  autoplot(turk_econ)
```

```{r}
s <- accuracy(fit) |> pull(RMSE)
yhat <- forecast(fit, h = 1) |> pull(.mean)
# SES
yhat[1] + c(-1, 1) * qnorm(0.975) * s[1]
## Holt
yhat[2] + c(-1, 1) * qnorm(0.975) * s[2]
```

```{r}
fit |>
forecast(h = 1) |>
mutate(PI = hilo(Exports, level = 95))
```

