---
title: "Lecture 4 Notes"
author: You!!
---

## Activity 1: Export Price of Salmon (Example 3.1)

```{r}
#| label: activity-1-template
library(astsa)

summary(fit <- lm(salmon~time(salmon), na.action=NULL))
tsplot(salmon, col=4, ylab="USD per KG", main="Salmon Export Price")
abline(fit)
```

1.  Does this time series appear stationary?

    no

2.  The (mathematical) equation in the book for the trend line above is:

    $$
    x_t = \beta_0 + \beta_1z_t + w_t, z_t = 2003\frac{8}{12}, 2003\frac{9}{12}, \dots, 2017\frac{5}{12}
    $$

    -   There is a typo in this equation. Correct the typo. (hint: examine to the first few entries of `time(salmon)`

        2001 –\> 2003 and 8/12 –\> 9/12

    -   Why are there fractions of the year? Explain what the fractional values mean and describe how they appear in the data set within R.

        They represent the month, i.e. 8/12 is Sep because it starts at 0 for Jan.

3.  Interpret the estimate of the slope.

    For each increase of 1-year, the average export price of salmon per kg increases by \$0.2529.

    For each increase of 1-month, the average export price of salmon per kg increases by \$0.2529 / 12.

## Activity 2: Trend Stationarity (Example 2.19)

Consider the time series model $$x_t = \beta t + y_t$$ Assume $y_t$ is stationary with mean function $\mu_y$ and and autocovariance function $\gamma_y(h)$

-   Compare this equation to the regression equation in the last example.

    The equations are similar with x_t being the same, beta1 = beta, beta0 = 0, and wt = yt, and zt = t

-   What are the mean function and autocovariance function of $x_t$?

    mean function: $E(x_t) = E(\beta t + y_t) = \beta t + \mu_y$

    autocovariance function: (see solutions)$\gamma_x(h) = cov(x_t+h, x_t) = E[(x_s - \mu)(x_t - \mu)] = \gamma_y(h)$

## Simulations

Take notes on your observations here!

## Activity 3

-   Given the code to generate the plot with the trend line, how would you view the equation of the trend line?
-   Visualize the de-trended series
-   Compute the acf of the salmon series and the detrended series. What do you notice?

```{r}
# as in the text
fit = lm(salmon~time(salmon), na.action=NULL) # the regression
par(mfrow=c(2,1)) # plot transformed data
tsplot(salmon, main="salmon price")
abline(fit)

## your code here

```
