---
title: "Lecture 5 - Draft"
author: Julia Schedler
format: 
  html: 
    code-fold: show
slide-number: true
execute: 
  eval: false
scrollable: true
---

## [Activity 1: Detrending a commodity (Example 3.7)]{.r-fit-text} {.smaller}

-   Given the code to generate the plot with the trend line, how would you view the equation of the trend line?

```{r}
#| eval: true

library(astsa)

fit <- lm(salmon ~ time(salmon), na.action = NULL)
tsplot(salmon, col=4, ylab='USE per KG', main='Salmon Export Price')
abline(fit)
summary(fit)
```
-   Visualize the de-trended series. Does it appear stationary?

```{r}
#| eval: false

residuals <- resid(fit)
tsplot(residuals)
```
-   Compute the acf of the salmon series and the detrended series. What do you notice?
```{r}
#| eval: false 

acf(salmon)
acf(residuals)
```


## [Activity 2: "assuming we are able to estimate it"]{.r-fit-text}

-   Look at pages 37-41 of the textbook
-   what is "it" in this context? (what are we estimating?)
-   If this is review, where did you first see these ideas?
-   Put a dot on the math anxiety rating distribution on the back board


## [Activity 3: Simulating a random walk and then differencing it]{.r-fit-text}
1. Simulate a random walk with no drift and plot it. 

```{r}
#| eval: false

## Simulate random walk wo/ drift
w <- rnorm(100)
x <- cumsum(w)

tsplot(x, main="random walk", ylab="", col=4)
 clip(0, 200, 0, 80)
 abline(a=0, b=0, lty=2, col=4) # drift (change b to drift value)
lines(x, col=6)
 clip(0, 200, 0, 80)
 abline(h=0, col=6, lty=2)
```
2. Use the `diff` function to difference the simulated series. Plot the result.
```{r}
#| eval: false

diff_x <- diff(x)


tsplot(diff_x, main="random walk", ylab="", col=4)
 clip(0, 200, 0, 80)
 abline(a=0, b=0, lty=2, col=4) # drift
lines(x, col=6)
 clip(0, 200, 0, 80)
 abline(h=0, col=6, lty=2)
```
3. Does this series appear stationary? How do you know? 

yes, at least in the mean

4. Visualize the ACF of the differenced series. Does it look like white noise?
```{r}
#| eval: false

acf(x)
acf(diff_x)
```



## [Activity 4: Differencing Salmon Prices]{.r-fit-text}
1. Compute and plot the differenced salmon series. 

```{r}
#| eval: false

tsplot(diff(salmon))
```

2. Does the series appear stationary?

No, bit of pseudo-cyclic

3. Visualize the acf of the differenced series. Does it look like white noise?
```{r}
#| eval: true

acf(diff(salmon))
acf1(diff(salmon))
```

Since the value at 1 is so large, it is hiding the pattern, so try acf1 --> see pattern

## [Activity 5: Comparing Differencing and Detrending]{.r-fit-text}
Compare the Acfs of the differenced and detrended salmon series. What do you notice?
```{r}
#| eval: false

acf(residuals)
acf(diff(salmon))
```

The differenced acf looks more like white noise then the detrended

SEE HERS FOR BETTER NOTES


detrending: x_t - hat_mu_t
differencing: x_t - x_t-1



