---
title: Lecture 16 Case Study
author: Julia Schedler
format: html
execute:
  echo: true
code-fold: show
slide-number: true
scrollable: true
---

Case Study

Consider the Cement production from the aus_production data set.

```{r}
#| message: false

library(tidyverse)
library(astsa)
library(fpp3)
```

```{r}
cement <- aus_production %>%
  select(Quarter, Cement)
```

Activity 1: Preliminary analysis

•	Produce a time series plot of the cement data

•	What are the beginning and ending dates for the time series?

•	What are the units for the cement production?

•	Propose an ETS model and an (S)ARIMA model for the data only based on the time series plot

```{r}
tsplot(as.ts(cement))

fits <- cement %>%
  model(
    ets_model = ETS(Cement ~ error("M") + trend("Ad") + season("M")),
    sarima_model = ARIMA(Cement ~ pdq(0, 0, 0) + PDQ(0, 0, 0)) 
    # Cant tell from plot alone, but definitely an increasing trend for a seasonal model
)
```

Activity 2: (S)ARIMA model

-   Follow the flow chart for “choose the model order yourself”

-   Identify p, d, q, P, D, Q, S as well as any transformations needed

```{r}
# tsplot(diff(box_cox(as.ts(cement), 0), 4))
# tsplot(diff(diff(box_cox(as.ts(cement), 0), 4)))
# tsplot(diff(diff(diff(box_cox(as.ts(cement), 0), 4))))

# trans_cement <- diff(diff(box_cox(as.ts(cement), 0), 4))
# trans_cement_2 <- diff(diff(diff(box_cox(as.ts(cement), 0), 4)))

# tells seasonal
box_cox(as.ts(cement), 0) %>% 
  as_tsibble() %>% 
  features(value, unitroot_nsdiffs)

tsplot(diff(box_cox(as.ts(cement), 0), 4))

# tells non-seasonal
diff(box_cox(as.ts(cement), 0), 4) %>% 
  as_tsibble() %>% 
  features(value, unitroot_ndiffs)

trans_cement <- diff(box_cox(as.ts(cement), 0), 4)

par(mfrow = c(1,2))
acf(trans_cement)
pacf(trans_cement)
```

|      | AR(p)                | MA(q)                | ARMA(p,q) |
|------|----------------------|----------------------|-----------|
| ACF  | Tails off            | Cuts off after lag q | Tails off |
| PACF | Cuts off after lag p | Tails off            | Tails off |

Non-seasonal
ACF: tailing off
PACF: cutting off after lag 1

Seasonal
ACF: Neither... tailing off in pseudo-cyclic way --> tailing off
PACF: Spikes after 2 seasonal periods

- p: 1
- d: 0
- q: 0
- P: 2
- D: 1
- Q: 0

Activity 3: Fit models

-   In one call to the model function, fit the following:

    -   The ETS model you proposed

    -   The (S)ARIMA model you selected in activity 2

    -   An automatically chosen ETS model

    -   An automatically chosen SARIMA model

-   Compare the automatically chosen and manually chosen models in terms of order/terms.

    -   Are any of the models equivalent?

-   Examine the residuals of each model.

```{r}
fits <- cement %>%
  model(
    ets_model = ETS(log(Cement) ~ error("M") + trend("Ad") + season("M")),
    sarima_model = ARIMA(log(Cement) ~ pdq(1, 0, 0) + PDQ(2, 1, 0, period = 4)),
    auto_ets = ETS(log(Cement)),
    auto_sarima = ARIMA(log(Cement))
)

fits

gg_tsresiduals(fits %>% select("ets_model"))
gg_tsresiduals(fits %>% select("sarima_model"))
gg_tsresiduals(fits %>% select("auto_sarima"))
```

CAN ONly USE AIC TO COMPARE MODELS WITH THE SAME DIFFERENCING VALUES

Activity 4: Model selection

-   Fit the same models as activity 3 but with just the data prior to (including 2007) Q4.

-   Are the automatically chosen models the same?

-   Can you make a choice of ETS vs. ARIMA?

-   Examine the residuals of each model

```{r}
train_cement <- cement %>%
  filter(Quarter <= yearquarter("2007 Q4"))
test_cement <- cement %>%
  filter(Quarter > yearquarter("2007 Q4"))

fits <- train_cement %>%
  model(
    ets_model = ETS(log(Cement) ~ error("M") + trend("Ad") + season("M")),
    sarima_model = ARIMA(log(Cement) ~ pdq(1, 0, 0) + PDQ(2, 1, 0, period = 4)),
    auto_ets = ETS(log(Cement)),
    auto_sarima = ARIMA(log(Cement))
)

fits

# ??
fits %>%
  forecast(h = 8) %>%
  accuracy(cement)

gg_tsresiduals(fits %>% select("auto_ets"))
gg_tsresiduals(fits %>% select("auto_sarima"))
```

The auto ets changed but the auto sarima is the same.

Activity 5: Forecast

-   Forecast the remaining data (2008 Q1 and on) for each of the methods.

-   Which model do you think is best?

```{r}
fits %>%
  select("auto_ets", "auto_sarima") %>%
  forecast(test_cement) %>%
  autoplot(test_cement, level = F) +
  theme_bw()
```
